Assessing reliability of an online behavioural laterality battery: a pre-registered

### 2020-07-17 online_laterality_reliability.Rmd
reliability writeup draft updated by DB. Being actively revised! Do not edit until DB says!

Zoe V. J., Woodhead

@@

<!--- not sure how corresponding authorship works in OSCCI lab --->
*Correspondence regarding this article should be addressed to Adam J. Parker, Department of Experimental Psychology, University of Oxford, Anna Watts Building, Woodstock Road, Oxford, OX2 6GG. Email: adam.parker@psy.ox.ac.uk.

@@

Studies of cerebral lateralisation often involve participants completing a series of perceptual tasks under laboratory conditions. This has constrained the number of participants recruited in such studies. The current work, however, introduces an online laterality battery that can be used to test participants on a variety of tasks: a rhyme decision visual half-field task, a dichotic listening task, a chimeric faces task, and a finger tapping task to assess manual preference. To validate the battery and assess its test-retest reliability, 392 participants completed the battery twice. Around a third of participants were left-handers. Not only do we report hemispheric advantages in the expected direction for each task, we additionally report good test-retest for the dichotic listening, chimeric faces, and finger tapping tasks. We also examine the association between handedness and atypical lateralisation and report that left-handers are more likely to depart from typical lateralisation. Furthermore, we provide data which indicates that lateralisation across tasks likely stems from multiple independent sources. The implications of these findings are discussed.

@@

Hemispheric specialisation, or lateralisation, is a ubiquitous characteristic of the human brain, yet many questions remain unanswered as to its causes and consequences. Much of the ambiguity in the published literature may arise from unreliable measures of lateralisation and underpowered studies. Without well powered experiments that use reliable tasks, it becomes difficult to address many research questions regarding the relationship between laterality indices and their stability over time. With respect to the relationship between indices, unless a task is reliable, the absence of a correlation cannot be interpreted as evidence for independence of function (Bishop, 2017). The best way to determine a task’s reliability is to test it twice. Here, the variation between participants should be larger than the variation within subjects. Taking this into consideration, the current work introduces an online behavioural laterality battery for which we established the test-retest reliability using a sample of 392 participants. The behavioural tasks included were: a novel rhyme decision task, a dichotic listening task (Hugdahl et al., 2009), a chimeric faces task (Burt & Perret, 1997), and a novel finger tapping task. We subsequently use this battery to we examine two pre-registered questions: (1) do extreme left-handers show a higher rate of atypical laterality than right handers on each behavioural laterality task; (2) do the laterality indices produced on each measure correlate.

@@

Behavioural laterality tasks have long been used to establish hemispheric dominance for certain cognitive functions. Rather consistently, results from visual half-field and dichotic listening paradigms suggest that the left hemisphere plays the dominant role in language processing (Hugdhal, 2000; Ocklenburg & Güntürkün, 2018). Here we review how each paradigm has been implemented to study language lateralisation and each paradigms’ reliability.

Evidence from the visual half-field paradigm (see Bourne, 2006, for a review), strongly points towards advantageous processing for letter strings presented in the right, relative to the left, visual field. As information in the right visual field projects contralaterally and arrives directly at the language dominant left hemisphere, it is thought to reflect an absence in callosal transfer that is required for the processing of linguistic information initially arriving at the right hemisphere.

Despite wide reporting of a right visual field advantage at the population level when using visual half-field tasks, the paradigm is not without critique. In particular, the validity of visual half-field tasks has been questioned. Krach, Chen, and Hartje (2006) compared the laterality indices (LIs) produced during a visual half-field experiment with those produced using functional Transcranial Doppler Sonography (fTCD). The visual half-field experiment itself consisted of lexical decisions made towards abstract nouns presented in the left- and right visual-fields. During the experiment, the blood flow to left and right hemispheres of 58 participants was measured. In addition, blood flow was measured during a word generation task—the gold standard task for assessing hemispheric dominance for language. Despite showing the expected right visual field advantage on the visual half-field task, the behavioural LIs produced on the visual half-field paradigm did not correlate with the fTCD LIs produced during word generation (r= .18) or the visual half-field task itself (r -.004). The lack of validity here likely reflects the poor reliability of the visual field task as split-half reliability was far from optimal (RT spilt-half: r= .14; accuracy split-half: r= .20). Thus, with such low split-half reliability it is unclear whether the task was able to accurately measure language lateralisation to start with.

In recent years it has become clear that the reliability and validity of visual half-field tasks to identify the language-dominant hemisphere is largely dependent on several methodological aspects. For example, Hunter and Brysbaert (2008) noted stronger and more stable visual-field differences when stimuli were presented bilaterally relative to those where only one stimulus was presented either in the LVF or in the RVF (Boles, 1987, 1990, 1994; see also Iacoboni & Zaidel, 1996). Brysbaert and Hunter argued that this is the result of stimuli competing during bilateral presentation, which is easier when the stimulus is presented in the language dominant hemisphere. Reliability and validity also depend on there being a sufficient number of trials and brief stimulus presentation. When accounting for these methodological aspects, the validity of visual half-field tasks seems promising. For example, Hunter and Brysbaert (2008) reported that LIs from visual half-field paradigms correlated positively with those from functional Magnetic Resonance Imaging (fMRI): word naming (r= .63), picture naming (r= .77). Here 192 trials were used during word naming and 160 trials during picture naming while stimuli were bilaterally presented and masked at offset. Furthermore, a promising visual half-field paradigm developed by Hausmann and colleagues (Hausmann et al., 2019; Willemin et al., 2016) which adhered to several of the methodological aspects outline by Brysbaert and Hunter has reported a right visual-field advantage at the population level with little variation occurring between languages. The authors, however, have not yet reported reliability measures for this task.

While the visual half-field technique measures lateralisation for visual stimuli, dichotic listening tasks examine hemispheric dominance for the processing of auditory stimuli. Like the visual half-field technique, the dichotic listening task makes use of the crossing of fibres at the corpus callosum. In this task, different streams of auditory linguistic stimuli (i.e. words, syllables) are presented to each ear and participants are either asked simply to report as many words as they can or to attend to one ear selectively. In either case, a right ear advantage has been demonstrated; participants report more verbal stimuli presented to the right ear (Hugdahl, 2011; Kimura, 1961). Again, this advantage has been attributed to stimuli being projected to the language dominant hemisphere without callosal transfer.

Over the years, various dichotic listening paradigms have been used (Bryden, 1988; Westerhausen, 2019). As with the visual half-field paradigm, the test-retest reliability of dichotic listening paradigms has been far from optimal (Voyer, 1998; Kelley & Littenberg, 2019). However, recent data from Westerhausen and Samuelsen (2020) indicated that when designed well, the dichotic listening paradigm can be reliable. First, based on a review of the published literature, Westerhausen and Samuelsen outlined eight design features that should be used in an optimal dichotic listening paradigm and explain their justification for each. The criteria are as follows: (1) stop consonant-vowel (CV) syllables as stimulus materials, (2) pair only CV stimuli from the same voicing category, (3) Alternating trials of voiced and unvoiced stimulus pairs, (4) free-recall instruction, (5) single-stimulus pair per trial, (6) paradigm length of 120 dichotic trials, (7) all stimulus pairs are presented in both orientations in identical frequency, and (8) includes binaural (homonymic) trials. Putting these criteria into practice, Westerhausen and Samuelsen administered a modified dichotic listening paradigm using both a verbal and manual response format to 50 right-handers. Interestingly, the full paradigm yielded a test-retest reliability of r= .93 for verbal responses and r= .91 for manual responses. Thus, the dichotic listening task provides a promising paradigm for those wanting to investigate language lateralisation without reliance on visual stimuli.

So far, the studies reviewed have been interested in language lateralisation. That is not to say that behavioural tasks are limited to linguistic materials. For instance, the visual half-field paradigm has been used to measure lateralisation for the processing of facial similarity (e.g. Gilbert & Bakan, 1973), face emotionality (e.g. Levy, Heller, Banich, & Burton, 1983), and local information (Sergent, 1982). In a recent review of the reproducibility of visual-field asymmetries, Brederoo, Nieuwenstein, Cornelissen, and Lorist (2019) reported that they were able to replicate five out nine perceptual asymmetries using the visual half-field paradigm.

Of importance to the current work is the use of the visual half-field paradigm to measure the lateralisation of face emotionality, in particular the chimeric faces task. In this task participants are presented with vertically split chimeric faces, where each half displays a different emotion. Depending on how the task is implemented, participants will either view a single chimeric face and report the emotion being expressed (e.g. Drebing, Federman, Edington, & Terzian, 1997; Kucharska-Pieture, & David, 2003) or view two faces, one above the other, and decide which of the two faces looks more emotive (e.g., Bourne, 2005; Levy, Heller, Banich, & Burton, 1983). In the two face version images will be identical, mirror images and typically include a neutral and emotive hemiface (i.e. Happiness-Neutral vs Neutral-Happiness). Regardless of implementation, participants’ responses point towards a bias for the processing of emotion in the left visual field. In the single face version, participants will more often report the emotion shown in the left hemiface. In the two-face version, participants show a bias in selecting the face with an emotion depicted in the left hemiface.

In terms of its reliability, the chimeric faces tasks seem promising. In Levy et al.’s (1983) early work, they reported good split-half reliability for the two faces version of the chimeric faces task, where the left and right handers both showed a split-half reliability of .93. The validity of the chimeric faces task has been demonstrated using behavioural and fMRI data. In Yovel, Tambini, and Brandman’s (2008) study, participants performed a behavioural experiment where they had to recognize centrally presented chimeric faces, which presented different identities in the right- and left-visual-field. Interestingly, the behavioural measure significantly correlated with the LI calculated based on activation in the fusiform face area (r= .49). However, it is of importance to note the sample size used here was small, but at face value the results seem promising.

@@

The studies reviewed so far have all focused on lateralisation at the population level. That is not to say that lateralisation does not vary between individuals. One factor that has been associated with the degree of lateralisation is handedness. Neuroimaging data strongly indicates that these differences in lateralisation are the result of a group of left-handers who are right language dominant rather than indicating that this relationship operates on a continuum (Mazoyer et al., 2014). This increased departure from typical lateralisation in left-handers is not only present in neuroimaging studies, it is also reflected in behavioural tasks. Thus it will be important for our battery to replicate the findings regarding handedness and lateralisation summarised below.

With regards to visual half-field techniques for studying language lateralisation, there is evidence to suggest that the visual field advantage varies with handedness (Elias, Bulman-Fleming, & McManus, 1999; Nicholls, 1994, 1998; Weekes, Capetillo-Cuncliffe, Rayman, Iacoboni, & Zaidel, 1999). However, a relatively well powered study, involving 100 participants conducted by Willemin et al. (2016) reported no effect of handedness (when coded as left and right) on the visual field advantaged during lexical decision task. At first, this finding may seem odd. However, it may reflect underlying issues with the task rather than represent strict evidence for a null effect of handedness on visual half-field task. Here, Willemin et al. asked participants to respond via manual response. This contrasts several studies that have reported increased variability when responding aloud. Thus, a task that lacks some active speech production component may not yield such strong visual field advantages that enable provide enough variation to detect subtle differences in handedness (Hunter & Brysbaert, 2008). This, of course, is something that we kept in mind when developing the visual half-field paradigm used in the current study.

When using dichotic listening paradigms to study language lateralisation, the evidence is clear that non-right-handers differ in terms of their laterality. While a clear right ear (i.e. left hemisphere) advantage is noted at the population level (see Hugdahl et al., 2009, and Westerhausen, 2019, for a review), there is evidence to suggest that this is reduced in left- relative to right-handers (Bethmann, Tempelmann, De Bleser, Scheich, & Brechmann, 2007; Bryden, 1970, 1988; Hugdahl, 2003); although handedness alone does not predict this difference (Van der Haegen, Westerhausen, Hugdahl, & Brysbaert, 2013). Across two studies Karlsson, Johnstone, and Carey (2019) provided a clear, yet comprehensive explanation of how performance on the dichotic listening task differed as a function of hand preference. Here, Karlsson et al. employed a consonant-vowel version of the dichotic listening task where participants reported which one of six syllables that they heard the clearest. For study 1, a clear right ear advantage was reported at the population level. Furthermore, right-handers were on average more lateralised than non-right-handers—87% of right-handers and 78% of non-right-handers showed a right ear advantage. For study 2, a similar pattern of results emerged. Again, right-handers were more lateralised with 85% of right-handers and 79% of non-right-handers showing an advantage. Together these results indicate that non-right-handers show a larger departure from the ‘typical’ lateralisation during dichotic listening tasks.

Recall that, for the chimeric faces, there is typically a left visual field advantage at the population level— indicating right hemisphere dominance. As with other laterality tasks, there appears to be a greater departure from this typical pattern of laterality for left-handers (e.g. Gilbert & Bakan, 1973; Heller & Levy, 1983; Karlsson et al., 2019; Levy et al., 1983; Rozkowski & Snelbecker, 1982). Again, this departure is well described by Karlsson et al. (2019). This departure was examined for two version of the chimeric faces task. The first involve participants viewing a single face with a different emotion shown in either hemiface. Participants then reported the emotion that they felt was most strongly displayed. The second was a two faces version where participants viewed two a face with an emotion on one side and the mirror version. Here participants had to decide which of the two faces showed a target emotion. For the first task, right-handers showed a bias for reporting information presented in the left visual field (i.e. right hemisphere); however, left-handers did not show a population bias. Specifically, 75% of right-handers and 60% of non-right-handers showed a leftwards bias. For the second chimeric faces task, a similar pattern emerged. 73% of right-handers and 57% of non-right-handers here showed a leftwards bias. Based on the above finding, a good laterality task must not only be reliable, but should be able to replicate a departure from typical lateralisation in those who are left-handed.

From the above reviewed studies, it appears that handedness is related to laterality across a range of tasks. This relationship points towards a general laterality factor which governs lateralisation across tasks and behaviours. This has been argued in several hypotheses (Behrmann & Plaut, 2015; Cook, 1984; Hellige, 1983; Kosslyn, 1987). Common to all hypotheses is the assumption that there would be a correlation between the degree of laterality across complementary tasks. While studies do report significant correlations between measures of lateralisation, these are often small (Bradzakova-Trajkov, Haberling, Roberts, & Corballis, 2010). Furthermore, there are a large number of studies which report no correlation between laterality indices (Bryden, Hecaen, & Deagostini, 1983; Rosch, Bishop, & Badcock, 2012; Whitehouse & Bishop, 2009; van der Haegen & Brysbaert, 2018). This has led to the conclusion that there are multiple independent biases across tasks which lead to hemispheric lateralisation of different cognitive domains.

@@

Our principle goal for the current study was to maximise the test-retest reliability of an online laterality battery. Specifically, we examined whether it is possible to achieve a test-retest reliability of .65 or above using online behaviour research methods, and, if so, whether this reliability could be achieved more efficiently by using fewer trials in each test. The tasks included in the battery were a novel rhyme decision visual half-field tasks, a consonant-vowel dichotic listening tasks, a chimeric faces visual half-field task, and a novel finger tapping tasks. As we were primarily interested in establishing test-retest reliability, the behavioural tasks included in the battery were completed twice on two different occasions. In addition to establishing the test-retest reliability for the battery, we tested several predictions about the relationship between laterality on the four behavioural tasks. We describe these predictions below.

Our first set of predictions related to participants’ performance at a population level on each task. Based on previous work, we predicted that: (1) the laterality indices for Rhyme Decision will show a right visual field advantage (indicating left hemisphere dominance for language processing); (2) the laterality indices for the Dichotic Listening task will show a right ear advantage (indicating left hemisphere dominance for language); (3) the laterality indices for Chimeric Faces will show a left visual field advantage (indicating right hemisphere dominance for face or emotion processing); and (4) the Finger Tapping task will show a right-hand advantage.

Our second prediction that we examined relates to handedness and atypical lateralisation. We predicted that strong left-handers (i.e. those with laterality quotient of -90 or less on the Edinburgh Handedness Inventory) will be more likely to show atypical laterality (i.e. laterality in the opposite direction to our population level predictions) than right-handers. This prediction follows from observations by Mazoyer et al (2014, p. 1) who studied a large group of left- and right-handers using fMRI to measure language laterality. They concluded “concordance of hemispheric dominance for hand and for language occurs barely above the chance level, except in a group of rare individuals (less than 1% in the general population) who exhibit strong right hemisphere dominance for both language and their preferred hand”.

Our final prediction related to the inter-relationships between laterality indices. With online testing, there is the opportunity to test inter-relationships between laterality indices with an adequately powered sample using measures of known reliability. Prior studies have been inconclusive because it has been difficult gather large amounts of data to establish test-retest reliability. For instance, Van der Haegen and Brysbaert (2018), found that language laterality did not correlate with face laterality in visual half-field tasks. However, their study was restricted to left-handers, who may be atypical in this regard, and they noted that lack of association was hard to interpret because it may be due to unreliability of measurement. Originally, we predicted that for right-handers, a model (a) where the two language tasks are correlated with each other but not with the chimeric faces will provide better fit to the data than a model (b) in which laterality indices are correlated between the language and chimeric face tasks, or a base model (c) where each task is independently lateralised. For left-handers, we anticipated that that model (c) may be a better fit to the data, indicating that laterality is dissociated across the different tasks. However, we were not able to test this planned analysis with the current data set as one task had low reliability. Thus, we instead examined the relationship between laterality tasks using correlation analysis.

@@

This project was pre-registered on the Open Science Framework (OSF) prior to the commencement of data collection. The registration form along with the analysis scripts and anonymised data can be found on the OSF at https://osf.io/qsxcv/.

@@

As our ultimate goal is to use these measures in individual differences research, we aimed to optimise the task to achieve observed test-retest reliability of at least .65. To achieve sufficient precision around the estimate of reliability, we estimated that we would require 300 participants. Out of the 300 participants, we originally specified that 80 would be left-handers. We had examined the power that a sample of 80 participants would provide to make the originally planned model comparisons outlined under hypothesis 2. By comparing model fit using a likelihood test, 80 participants would provide 100% power for model (a) and 97% power for model (b). Thus, we would have sufficient power to assess the fit of each model using Akaike weights, which are more conservative than likelihood tests.

Despite setting out to recruit 300 participants, our final sample exceeded this target (we say more about this below). In total, 441 native English speakers were initially recruited through Prolific Academic. Participants had normal or corrected-to-normal vision and reported no history of neurological disease (e.g. head injury, brain tumour, stroke) or significant hearing loss/impairment.

@@

#### Test of ocular dominance

Adapted versions of the Miles test (Miles, 1929) and the Porta test (Porac & Coren, 1976) were used to determine each participant’s eye dominance in central gaze (i.e., when looking straight ahead). Each test classified participants as being either left or right eye dominant.

#### Edinburgh handedness inventory

The Edinburgh Handedness Inventory (Oldfield, 1971) was administered to quantify handedness on a continuum. For 10 everyday activities, participants indicated their preferred hand use on a 5-step scale (right hand strongly preferred, right hand preferred, no preference, left hand preferred, left hand strongly preferred). The Edinburgh handedness inventory was scored in the standard way, which involves combining information about direction and exclusiveness of hand preference. For simple contrast between left- and right-handers, we treated indices greater than 0 as right handed, and indices less than 0 as left handed. For analysis of extreme left-handers, we focussed on left-handers with an index of -90 or less, which corresponds to near-exclusive use of the left hand.

>>>>

Zoe V. J. Woodhead

@@

<!--- not sure how corresponding authorship works in OSCCI lab --->
<!--- I'm fine with you in this role--->
*Correspondence regarding this article should be addressed to Adam J. Parker, Department of Experimental Psychology, University of Oxford, Anna Watts Building, Woodstock Road, Oxford, OX2 6GG. Email: adam.parker@psy.ox.ac.uk.

@@

Studies of cerebral lateralisation often involve participants completing a series of perceptual tasks under laboratory conditions. This has constrained the number of participants recruited in such studies. Online testing can potentially allow for much larger sample sizes, but limits the amount of experimental control that is feasible. Here we considered whether online testing could give valid and reliable results on four tasks: a rhyme decision visual half-field task, a dichotic listening task, a chimeric faces task, and a finger tapping task to assess manual preference. We recruited 392 participants, oversampling left-handers, who completed the battery twice. Three of the tasks showed evidence of both validity and reliability, insofar as they showed hemispheric advantages in the expected direction and test-retest reliability of at least .65. The rhyme judgement task was less satisfactory. We also found that left-handers were more likely to depart from typical lateralisation. Contrary to prediction, lateralisation across tasks was not significantly correlated. We conclude that three of the tasks, dichotic listening, chimeric faces and finger-tapping, show considerable promise for online evaluation of cerebral lateralisation, but the results do not support the notion of a common underlying cause of lateralisation for different cognitive tasks.

@@

Hemispheric specialisation, or lateralisation, is a ubiquitous characteristic of the human brain, yet many questions remain unanswered as to its causes and consequences. Much of the ambiguity in the published literature may arise from unreliable measures of lateralisation and underpowered studies. Without well-powered experiments that use reliable tasks, it becomes difficult to address many research questions regarding the relationship between laterality indices and their stability over time. With respect to the relationship between indices, unless a task is reliable, the absence of a correlation cannot be interpreted as evidence for independence of function (Bishop, 2017). Reliability can either be assessed within a test session (e.g. split-half reliability, from correlations between odd and even items) or from repeated assessments on different occasions, test-retest reliability. The latter is generally preferred as it includes all test items and will incorporate fluctuations in performance due to changes in state or test conditiions from one occasion to another. Assessment of test-retest reliability, however, requires large samples of people who undergo repeated testing, and this is seldom feasible for in-person testing. Online testing allows for recruitment of much larger samples, but raises questions about test validity when opportunities for experimental control are reduced. Here we introduce an online behavioural laterality battery for which we established the test-retest reliability using a sample of 392 participants. The behavioural tasks included were: a novel rhyme decision task, a dichotic listening task (Hugdahl et al., 2009), a chimeric faces task (Burt & Perret, 1997), and a novel finger tapping task. Validity of the tasks was assessed in terms of whether lateralisation was observed in the predicted direction, based on prior literature with in-person testing on analogous tasks. We also examined two pre-registered questions: (1) do extreme left-handers show a higher rate of atypical laterality than right handers on each behavioural laterality task? (2) do the laterality indices from different tasks correlate?

@@

Behavioural laterality tasks have long been used to establish hemispheric dominance for certain cognitive functions. Rather consistently, results from visual half-field and dichotic listening paradigms support the view that the left hemisphere plays the dominant role in language processing (Hugdahl, 2000; Ocklenburg & Güntürkün, 2018). Here we review how each paradigm has been implemented to study language lateralisation and what is known about reliability and validity.

#### Visual half-field paradigm
In this paradigm, visual stimuli are briefly presented to the left or right of a central fixation stimulus. The visual pathways from the left and right sides of space to visual cortex are crossed for non-foveal viewing, and so the right eye projects directly to the left hemisphere, and the left eye to the right hemisphere. The corpus callosum allows for transfer between hemispheres, but the rationale of the method is that there will be a processing advantage for language stimuli that are directly projected to the language-dominant hemisphere. This method has been used since the 1950s (Mishkin & Forgays, 1952). Bourne (2006) reviewed the literature, which consistently shows advantageous processing for letter strings presented in the right, relative to the left, visual field in most people.

Despite wide reporting of a right visual field advantage at the population level,  the validity of visual half-field tasks has been questioned. Krach, Chen, and Hartje (2006) compared the laterality indices (LIs) produced during a visual half-field experiment with those produced using functional Transcranial Doppler Sonography (fTCD). The visual half-field task consisted of lexical decisions made towards abstract nouns presented in the left- and right visual-fields. During the experiment, the blood flow to left and right hemispheres of 58 participants was measured. In addition, blood flow was measured during a word generation task — the gold standard task for assessing hemispheric dominance for language. Despite showing the expected right visual field advantage on the visual half-field task, the behavioural LIs produced on the visual half-field paradigm did not correlate with the fTCD LIs produced during word generation (r= .18) or the visual half-field task itself (r -.004). The lack of validity here likely reflects the poor reliability of the visual field task as split-half reliability was very low (RT spilt-half: r= .14; accuracy split-half: r= .20). This illustrates the difficulty of drawing conclusions about associations between laterality indices when reliability of measurement is so low.

In recent years it has become clear that the reliability and validity of visual half-field tasks to identify the language-dominant hemisphere is largely dependent on several methodological aspects. For example, Hunter and Brysbaert (2008) noted stronger and more stable visual-field differences when stimuli were presented bilaterally relative to those where only one stimulus was presented either in the LVF or in the RVF (Boles, 1987, 1990, 1994; see also Iacoboni & Zaidel, 1996). Brysbaert and Hunter argued that this is the result of stimuli competing during bilateral presentation, enhancing the advantage to the stimulus presented in the language dominant hemisphere. Reliability and validity also depend on there being a sufficient number of trials and brief stimulus presentation to prevent eye movements. When accounting for these methodological aspects, the validity of visual half-field tasks seems promising. For example, Hunter and Brysbaert (2008) used two visual half-field tasks, word naming (192 trials) and picture naming (160) in which stimuli were bilaterally presented and masked at offset. The two tasks correlated .8 (95% CI .598-.956) in a sample of 26 left-handers. Encouragingly, the LIs from these visual half-field tasks also correlated positively with laterality from functional Magnetic Resonance Imaging (fMRI) on a word generation task in 10 left-handers. However, the small sample size means that the confidence intervals around these estimates are very wide: word naming, r= .63 (95% CI = .001-.902), picture naming, r= .77 (95% CI = .272-.943).  <!---I removed Hausmann et al in interests of length, as they didn't report reliability--->

#### Dichotic listening
Simultaneous presentation of different auditory stimuli to left and right ears for immediate recall was a method originally used in studies of attention, but it became evident that there was usually an advantage for verbal stimuli presented to the right ear (Kimura, 1967). Both ears project to both hemispheres, but the weaker ipsilateral pathways appear to be suppressed when the two sides are in competition, and so the right ear, with a contralateral projection to the left hemisphere, will have an advantage in tasks that involve language processing.

Over the years, various dichotic listening paradigms have been used (Bryden, 1988; Westerhausen, 2019). As with the visual half-field paradigm, the test-retest reliability of dichotic listening paradigms has been far from optimal (Voyer, 1998; Kelley & Littenberg, 2019). However, recent data from Westerhausen and Samuelsen (2020) indicated that when designed well, the dichotic listening paradigm can be reliable. First, based on a review of the published literature, Westerhausen and Samuelsen outlined eight design features that should be used in an optimal dichotic listening paradigm and explain their justification for each. The criteria are as follows: (1) stop consonant-vowel (CV) syllables as stimulus materials, (2) pair only CV stimuli from the same voicing category, (3) Alternating trials of voiced and unvoiced stimulus pairs, (4) free-recall instruction, (5) single-stimulus pair per trial, (6) paradigm length of 120 dichotic trials, (7) all stimulus pairs are presented to the two sides in identical frequency, and (8) includes binaural (homonymic) trials. Putting these criteria into practice, Westerhausen and Samuelsen administered a modified dichotic listening paradigm using both a verbal and manual response format to 50 right-handers. Interestingly, the full paradigm yielded a test-retest reliability of r= .93 for verbal responses and r= .91 for manual responses. Thus, the dichotic listening task provides a promising paradigm for those wanting to investigate language lateralisation without reliance on visual stimuli.

#### Chimeric faces

In studies of face processing, it is possible to use a modification of the visual half-field paradigm in which single stimuli are viewed, composed of vertically split chimeric faces, where each half displays a different emotion or identity. Our focus here is on emotion detection in chimeric faces, where a bias is found towards reporting the emotion shown in the left hemiface.

The reliability of the chimeric faces tasks seem promising. Early work by Levy et al (1983) using two vertically presented chimeric faces, found split-half reliability of .93. The validity of the chimeric faces task has been demonstrated using behavioural and fMRI data, but in a task of recognising facial identity rather than emotion. In Yovel, Tambini, and Brandman’s (2008) study, participants performed a behavioural experiment where they had to recognize centrally presented chimeric faces, which presented different identities in the right- and left-visual-field. It is encouraging to note that the behavioural measure significantly correlated with the LI calculated based on activation in the fusiform face area (r= .49). However, the sample size used here was small: with 17 participants, the 95% confidence interval around the observed correlation of .49 ranged from .012 to .786.

@@

The studies reviewed so far have indicated consistent population biases in lateralisation on behavioural tasks, but also individual differences. One factor that has been associated with the degree of lateralisation is handedness. Neuroimaging data indicates that for language tasks these differences in lateralisation reflect the contribution of a subset of strong left-handers who are right language dominant, rather than indicating that this relationship operates on a continuum (Mazoyer et al., 2014). This association can be used to validate behavioural measures of lateralisation: we would expect to see departure from typical lateralisation in strong left-handers.

With regards to visual half-field techniques for studying language lateralisation, there is evidence to suggest that the visual field advantage varies with handedness (Elias, Bulman-Fleming, & McManus, 1999; Nicholls, 1994, 1998; Weekes, Capetillo-Cuncliffe, Rayman, Iacoboni, & Zaidel, 1999). However, a relatively well powered study, involving 100 participants conducted by Willemin et al. (2016) reported no hint of any effect of handedness, when coded as left and right, or when divided into left (N = 19), mixed (N = 24) and right-handed (N = 57) groups, on the visual field advantage during a lexical decision task. This result illustrates the recurring challenge of interpreting null results in this field: do they reflect a true lack of relationship, poor test reliability, inadequate power to detect small effects, or specific methodological factors? We may note that Willemin et al. asked participants to respond via manual response, whereas other studies have involved an active speech production component, which could influence task lateralisation.

On dichotic listening paradigms, the clear right ear (i.e. left hemisphere) advantage that is noted at the population level (see Hugdahl et al., 2009, and Westerhausen, 2019, for a review) is reduced in left-handers (Bethmann, Tempelmann, De Bleser, Scheich, & Brechmann, 2007; Bryden, 1970, 1988; Hugdahl, 2003). More recently, Karlsson, Johnstone, and Carey (2019) employed a consonant-vowel version of the dichotic listening task where participants reported which one of six syllables that they heard the clearest. In study 1, 87% of right-handers and 78% of non-right-handers showed a right ear advantage. For study 2, 85% of right-handers and 79% of non-right-handers showed a right ear advantage.

For chimeric faces, which typically show a left visual field advantage there again appears to be a greater departure from the typical pattern of laterality for left-handers (e.g. Gilbert & Bakan, 1973; Heller & Levy, 1983; Levy et al., 1983; Rozkowski & Snelbecker, 1982). This was also studied in the study by Karlsson et al. (2019). When viewing a single face with a different emotion shown in either hemiface, 75% of right-handers and 60% of non-right-handers showed a leftwards bias. For the second chimeric faces task that involved two chimeric faces arranged vertically, a similar pattern emerged. 73% of right-handers and 57% of non-right-handers showing a leftwards bias.

From the above reviewed studies, it appears that handedness is related to laterality across a range of tasks. This relationship points towards a general laterality factor which governs lateralisation across tasks and behaviours. This has been argued in several hypotheses (Behrmann & Plaut, 2015; Cook, 1984; Hellige, 1983; Kosslyn, 1987). Common to all hypotheses is the assumption that there would be a correlation between the degree of laterality across complementary tasks. While studies do report significant correlations between measures of lateralisation, these are often small (Bradzakova-Trajkov, Haberling, Roberts, & Corballis, 2010). Furthermore, there are a large number of studies which report no correlation between laterality indices (Bryden, Hecaen, & Deagostini, 1983; Rosch, Bishop, & Badcock, 2012; Whitehouse & Bishop, 2009; van der Haegen & Brysbaert, 2018). However, it has not been possible to clearly unpick whether the null results indicate that there are multiple independent lateral biases across tasks, or whether there is a more prosaic explanation in terms of poor reliability or validity of tasks.

@@

The principal goal of the current study was to evaluate the reliability and validity of an online behavioural laterality battery. Specifically, we examined whether it is possible to observe the usual lateral biases at the population level in tests of language and emotion processing, and whether it is possible to achieve a test-retest reliability of .65 or above using online behaviour research methods. In addition, for tasks where satisfactory reliability was achieved, we considered how many trials were needed to achieve this. The tasks included in the battery were a novel rhyme decision visual half-field task, a consonant-vowel dichotic listening tasks, a chimeric faces visual half-field task, and a novel finger tapping task. These behavioural tasks were completed twice on two different occasions.

Our first set of predictions concerned validity of the tasks as indicators of cerebral lateralisation. Based on previous work, we predicted that at the population level: (1) the laterality indices for Rhyme Decision would show a right visual field advantage (indicating left hemisphere dominance for language processing); (2) the laterality indices for the Dichotic Listening task would show a right ear advantage (indicating left hemisphere dominance for language); and (3) the laterality indices for Chimeric Faces would show a left visual field advantage (indicating right hemisphere dominance for face or emotion processing). The Finger Tapping task was designed to provide a quantitive measure of degree of handedness, and was expected to show a right-hand advantage in the overall sample.

Our second prediction related to handedness and atypical lateralisation. We predicted that strong left-handers (i.e. those with laterality quotient of -90 or less on the Edinburgh Handedness Inventory) would be more likely to show atypical laterality (i.e. laterality in the opposite direction to our population level predictions) than right-handers. This prediction followed from observations by Mazoyer et al (2014, p. 1) who studied a large group of left- and right-handers using fMRI to measure language laterality. They concluded “concordance of hemispheric dominance for hand and for language occurs barely above the chance level, except in a group of rare individuals (less than 1% in the general population) who exhibit strong right hemisphere dominance for both language and their preferred hand”.

Our final prediction related to the inter-relationships between laterality indices. With online testing, there is the opportunity to test inter-relationships between laterality indices with an adequately powered sample using measures of known reliability. Prior studies have been inconclusive because it has been difficult gather large amounts of data to establish test-retest reliability. For instance, Van der Haegen and Brysbaert (2018), found that language laterality did not correlate with face laterality in visual half-field tasks. However, their study was restricted to left-handers, who may be atypical in this regard, and they noted that lack of association was hard to interpret because it may be due to unreliability of measurement. In the preregistration, we predicted that for right-handers, a model (a) where the two language tasks are correlated with each other but not with the chimeric faces will provide better fit to the data than a model (b) in which laterality indices are correlated between the language and chimeric face tasks, or a base model (c) where each task is independently lateralised. For left-handers, we anticipated that that model (c) may be a better fit to the data, indicating that laterality is dissociated across the different tasks. However, we were not able to test this planned analysis with the current data set as one language task had low reliability. Thus, we confined our analysis to presentation of correlations between laterality indices from the four measures.

@@

This project was pre-registered on the Open Science Framework (OSF) prior to the start of data collection. The registration form along with the analysis scripts and anonymised data can be found on the OSF at https://osf.io/qsxcv/.

@@

As our ultimate goal is to use these measures in individual differences research, we aimed to optimise the task to achieve observed test-retest reliability of at least .65. Our initial power analysis showed that to achieve sufficient precision around the estimate of reliability, we would require 300 participants. Out of the 300 participants, we originally specified that 80 would be left-handers. This would give us power to apply our planned model-fitting analysis to left-handers as a separate group.

In practice, our final sample exceeded 300, because we discarded participants from individual tasks for failing to meet prespecified criteria (see below). In total, 441 native English speakers were initially recruited through Prolific Academic. Participants had normal or corrected-to-normal vision and reported no history of neurological disease (e.g. head injury, brain tumour, stroke) or significant hearing loss/impairment.

@@

#### Tests of ocular dominance

Adapted versions of the Miles test (Miles, 1929) and the Porta test (Porac & Coren, 1976) were used to determine each participant’s eye dominance in central gaze (i.e., when looking straight ahead). Each test classified participants as being either left or right eye dominant. In practice, the two tests agreed perfectly, so we report here just the Miles test.

#### Edinburgh Handedness Inventory

The Edinburgh Handedness Inventory (EHI: Oldfield, 1971) was administered to quantify handedness on a continuum. For 10 everyday activities, participants indicated their preferred hand use on a 5-step scale (right hand strongly preferred, right hand preferred, no preference, left hand preferred, left hand strongly preferred). The EHI was scored in the standard way, which involves combining information about direction and exclusiveness of hand preference. For simple contrast between left- and right-handers, we treated indices greater than 0 as right handed, and indices less than 0 as left handed. For analysis of extreme left-handers, we focussed on left-handers with an index of -90 or less, which corresponds to near-exclusive use of the left hand.
